The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?

### 2025-08-29 thesis/introduction.tex
julio edits at durham

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These little galaxies are typically
defined to be fainter than the LMC \citep[\(M_V \gtrsim -18\) or
similarly \(L \lesssim 10^9 L_\odot\),
e.g.,][]{hodge1971, mcconnachie2012}. As the smallest class of galaxies,
dwarfs are the most cosmologically numerous galaxies. Dwarf galaxies are
highly \emph{dark-matter dominated}, with mass to light ratios exceeding
100 to 1000 \(M_\odot/ L_\odot\). Because of their small total masses,
many dwarf galaxy satellites of the Milky Way are \emph{quenched}, with
little to no recent star formation. Milky Way dwarfs contain stellar
populations which are \emph{relics} from the early universe, consisting
of many of the oldest and most metal poor stars. Understanding the
properties of dwarf galaxies thus has implications across astronomy,
from cosmological structure formation on the smallest scale to extremely
metal-poor stellar populations to chemical evolution to galactic
dynamics. Of the classical dwarfs, we find that the Sculptor and Ursa
Minor dwarf spheroidal galaxies stand out with a more extended density
profile relative to an exponential, hinting at tidal effects. As a case
study in tidal effects and our understanding of dwarf galaxy formation
and evolution, we aim to understand the dynamical history of these
galaxies.

In this section, we first describe cosmological structure formation and
the role of dwarf galaxies. Then, we explore the observational history
of dwarf galaxies, the \emph{Gaia} telescope, open questions, and why
Sculptor and Ursa Minor stand out. We discuss idealized simulations,
tidal effects, break radii, and recent developments in tidal
simulations. Finally, we provide a roadmap to the thesis as a whole at
the end of this section.

\section{Observational context}\label{observational-context}

Dwarf galaxies have long raised conundrums in theories of galaxy
formation. The discovery of Fornax and Sculptor in 1938
\citep{shapley1938}\footnote{Technically, the Large and Small Magellanic
  Clouds (LMC, SMC) may be classified as dwarf galaxies, but these were
  likely always known to humans at southern latitudes.}, with no known
analogues, already presented such an enigma. Shapley presented these
dwarfs as a new type of \emph{stellar system} resembling the Magellanic
Clouds and globular clusters but did not attempt to speculate on the
exact nature. While dwarf galaxies were quickly understood to be
galaxies based on the inferred luminosities and sizes, their nature and
formation remained unclear for decades
\citep[e.g.,][]{hodge1971, gallagher+wyse1994}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial amounts of dark matter. From early determinations of the
velocity dispersion for the Sculptor and Ursa Minor dwarf spheroidal
(dSph) galaxies
\citep[e.g.,][\citet{aaronson+olszewski1987}]{aaronson1983}, inferred
mass-to-light ratios were at least 10 times larger than the values from
globular clusters scaled to the same sizes. While rather uncertain
initially, these values have corroborated with larger and more precise
samples \citep[e.g.,][]{hargreaves+1994}. Subsequently, several theories
attempting to understand the formation and observed properties of these
objects were proposed. Examples include: dwarf galaxies are undergoing
tidal dissolution resulting in extreme mass-to-light ratios
\citep[e.g.,][]{kuhn+miller1989}, presence of massive central black
holes \citep[e.g.,][]{strobel+lake1994}, the formation of dark
matter-free ``tidal dwarfs'' from past mergers
\citep[e.g.,][\citet{kroupa1997}]{lynden-bell1982}, or modified theories
of gravity \citep{milgrom1995}. However, consistency with CDM galaxy
formation was also not out of the question
\citep[e.g.,][]{dekel+silk1986}. Since then, we have known that dwarf
galaxies are among the darkest objects in the universe, and
understanding their properties is critical to understanding dark matter.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
and recent star formation. Fig.~\ref{fig:galaxy_images} shows the Large
Magellanic Cloud (LMC) as an example of an irregular, bright dwarf
galaxy displaying a bar. \textbf{Classical dwarfs} occupy intermediate
luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these
systems are old, gas-poor, and spheroidal. All Milky Way satellites
discovered before digital sky surveys are classicals. The 12 classical
dwarfs satellites of our Galaxy are Sagittarius, Fornax, Leo I,
Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
I, Sextans I, and Crater II.\footnote{While formally the dwarf galaxy
  names we discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor
  dSph, we omit this suffix for brevity.} The \textbf{ultra-faint}s
occupy the very faintest magnitudes (\(-7.7 \lesssim M_V\)). These
galaxies have minuscule stellar masses, tend to be more compact, and are
the darkest known galaxies. Altogether, dwarf galaxies span more than 15
orders in absolute magnitude.

Since Local Group dwarfs are nearby, we can study these galaxies on a
star-by-star basis. As a result, it is possible to measure the 3D
velocity and position of a star if we can measure the stars position,
line-of-sight (LOS) velocity, and proper motions. Unfortunately,
determining distances and full 3D velocities is challenging. The most
direct measurement of distance, parallax, requires precise tracking of a
star's sky position across a year. And while line-of-sight (LOS)
velocities are easily determined from spectroscopy, tangential
velocities, derived from proper motions and distances, are much more
challenging. Typically, measuring proper motions requires accurate (much
less than arcsecond) determinations of small changes in a star's
position over baselines of several years to decades. The full 6D
position and velocity information for stars has historically been known
for only a handful of stars.

@@

@@

\emph{(db: these paragraphs start the same\ldots)}

\emph{Gaia} has redefined astrometry, providing photometry, positions,
proper motions, and parallaxes for over 1 billion stars
\citep{gaiacollaboration+2021}. Launched in 2013, \emph{Gaia} is a
space-based, all-sky survey telescope situated at the Sun-Earth L2
Lagrange point \citep{gaiacollaboration+2016}. While \emph{Gaia}
completed its space-based mission in 2025, two forthcoming data releases
remain. Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imagining two regions
separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
tiny changes in relative positions of stars across small and large
angles. Combining measurements from multiple epochs across several
years, an absolute all-sky reference frame is derived from which
parallax and proper motions are derived. In addition to astrometry,
\emph{Gaia} measures photometry in the wide \emph{G} band (330-1050nm)
and colours from the blue photometer (BP, 330-680 nm) and red photometer
(RP, 640-1050 nm). \emph{Gaia} additionally provides low resolution
BP-RP spectra and radial velocity measurements of bright stars
(\emph{Gaia} radial velocity magnitudes \textless16)
\citep{gaiacollaboration+2016}. For this work, \emph{Gaia}'s most
relevant measurement are \(G\) magnitude, \(G_{\rm BP} - G_{\rm RP}\)
colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
  change in right ascension, \(\alpha\), and declination, \(\delta\),
  typically in units of mili-arcsecond (mas) per year.
  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
  effects in \(\alpha\).}

\emph{Gaia} has revolutionized many fields of astronomy including
studies of the Milky Way and the Local Group. The Local Group is defined
as galaxies which are approximately bound to the Milky Way-Andromeda
system, or within about 1 Mpc from the Milky Way \citep[e.g.,][ and
references therein]{mcconnachie2012}. For example, 6D kinematic
classification of Milky Way stars led to the discovery of past mergers
or Milky Way building blocks like \emph{Gaia}-Sausage Enceladus
\citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
\emph{Gaia} snail {[}e.g., \citet{antoja+2018}, and dynamical effects of
the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
helped find and constrain numerous stellar streams
\citep{bonaca+price-whelan2025}. Altogether, \emph{Gaia} has revealed
the hierarchical formation and complex, evolving structure of our own
Galaxy.

For Milky Way satellites, \emph{Gaia} has enabled well-constrained
orbital analysis and facilitated precise stellar membership
determination. Before, proper motions of dwarf galaxies were sparse or
undetermined. Few galaxies had precisely measured proper motions, often
from Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
\emph{Gaia} allowed for the first systematic determinations of Milky Way
satellite proper motions of unprecedented precision
\citep{MV2020a, pace+li2019}. While the proper motion uncertainty on a
typical dwarf member star is often large, by combining the proper
motions for 100s to 1000s of stars in \emph{Gaia}, precise proper motion
measurements can be determined, sometimes only limited by \emph{Gaia}'s
systematic error floor \citep[e.g.,][]{MV2020a}. Proper motions have
ushered in a new dynamical era for MW satellite studies, where we can
derive precise orbits assuming a given MW potential. In addition,
\emph{Gaia} helps separate out contaminating MW foreground stars. By
measuring parallaxes or proper motions, many background and foreground
stars can be removed as non-members
\citep[e.g.,][\citet{jensen+2024}]{battaglia+2022}.

@@

Today, we know that the Milky Way system is teeming with satellites.
Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
including dwarf galaxies, ambiguous systems (whose nature as a
dwarf-like or cluster-like system remains uncertain), and globular
clusters. Advances in observational techniques has accelerated progress
in our understanding of dwarf galaxies and introduced new questions.
Deep digital photometric surveys have more than quadrupled the number of
known Milky Way dwarf galaxies \citep{simon2019}. Upcoming and ongoing
surveys, like the Vera Rubin Observatory's Legacy Survey for Space and
Time \citep{ivezic+2019}, will likely identify even fainter dwarf
galaxies. In addition, large aperture multi-object spectrographs have
revealed the detailed and complex inner chemodynamical structure of
dwarf galaxies. Beyond precise structural and kinematic properties of
dwarf galaxies, modern observations allow for the separation of multiple
stellar populations, detailed constraints on the dark matter density
profiles, and hints of tidal disruption or stellar halos.

Of local dwarfs, the classical systems still remain the most studied
galaxies with the best constrained parameters. Because many Milky Way
dwarfs extend 1-2 degrees across the sky, measurements of some of their
properties can be challenging \citep[e.g.,][]{mateo1998}. These dwarfs
also contain large numbers of bright (giant) stars which can be followed
up spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a
result, the determination of fundamental properties such as the
position, size, orientation, distances, proper motions, line-of-sight
(LOS) velocities, and velocity dispersions \(\sigma_v\), are all
relatively well constrained today. However, ongoing research continues
to redefine our understanding of the detailed structure of Milky Way
satellites, hinting that these objects may be more complex than expected
at first glance.

Observations of dwarf galaxies have been the origin of several disputes
or \emph{small-scale} problems for \(\Lambda\)CDM \citep[see
review][]{bullock+boylan-kolchin2017}. For example, the mismatch between
the expected number of dwarf galaxies from simulations and the observed
abundance was known as the \emph{missing satellites problem}.
Additionally, a number of observations suggested that some dwarf
galaxies, although not all, possess dark matter ``cores''
\citep[e.g.,][]{moore1994, adams+2014, oh+2015, walker+penarrubia2011, read+walker+steger2019}.\footnote{In
  detail, gas-phase rotation curves are better able to differentiate
  between cores and cusps, whereas stellar kinematics is less
  constraining.} As a result, alternative forms of dark matter have been
advocated as solutions, such as Warm and Self-Interacting Dark Matter.

>>>>

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These galaxies are typically defined to be
fainter than the Large Magellanic Cloud (LMC), with \(M_V \gtrsim -18\)
or similarly \(M_\star \lesssim 10^9 M_\odot\) {[}e.g.,
\citet{hodge1971}; \citet{mcconnachie2012}). Because the galaxy
luminosity function increases towards fainter objects, dwarfs are the
most numerous galaxies in the Universe. Dwarf galaxies are also highly
\emph{dark-matter dominated}, with mass to light ratios which may exceed
1000 \(M_\odot/ L_\odot\). With the exception of the Magellanic Cloud,
all dwarf galaxy satellites of the Milky Way (MW) are \emph{quenched},
with little to no recent star formation. Indeed, most faint MW
satellites contain stellar populations which are \emph{relics} from the
early universe, consisting of many of the oldest and most metal poor
stars.

Understanding the properties of dwarf galaxies thus has implications
across astronomy, from cosmological structure formation on small scales
to the formation of metal-poor stellar populations.

\emph{Remove this paragraph?} Of the classical dwarfs, we find that the
Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
extended density profile relative to an exponential, hinting at tidal
effects. As a case study in tidal effects and our understanding of dwarf
galaxy formation and evolution, we aim to understand the dynamical
history of these galaxies.

In this section, we first describe the general observed properties of
local dwarf galaxies. Next, we summarize the cosmological origin and
role of dwarf galaxies. We then review recent updates to the observed
properties of dwarf galaxies made on new facilities. We then discuss how
tides influence satellite galaxies in general. Finally, we provide a
roadmap to the thesis. \textbf{Look at this paragraph again}

\section{Observational history of dwarf
galaxies}\label{observational-history-of-dwarf-galaxies}

Dwarf galaxies have long raised conundrums for theories of galaxy
formation. The discovery of Fornax and Sculptor in 1938
\citep{shapley1938}\footnote{Technically, the Large and Small Magellanic
  Clouds (LMC, SMC) are also classified as dwarf galaxies, but these
  were likely always known to humans at southern latitudes.}, with no
known analogues, already presented such an enigma. Shapley presented
these dwarfs as a new type of \emph{stellar system} resembling the
Magellanic Clouds and globular clusters but did not attempt to speculate
on their exact nature. While dwarf galaxies were quickly understood to
be galaxies based on the inferred luminosities and sizes, their nature
and formation remained unclear for decades
\citep[e.g.,][]{hodge1971, gallagher+wyse1994}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial amounts of dark matter. From early determinations of the
velocity dispersion for the Sculptor and Ursa Minor dwarf spheroidal
(dSph) galaxies \citep[e.g.,][]{aaronson1983, aaronson+olszewski1987},
inferred mass-to-light ratios were at least 10 times larger than the
values of globular clusters (GCs). While rather uncertain initially,
these values have been corroborated with larger and more precise samples
\citep[e.g.,][]{hargreaves+1994}. At the time, several theories were
proposed to explain these unusually high mass-to-light ratios. Examples
include: ongoing tidal disruption artificially inflating the
line-of-sight velocity dispersions \citep[e.g.,][]{kuhn+miller1989}, the
presence of massive central black holes
\citep[e.g.,][]{strobel+lake1994}, or modified theories of gravity
\citep{milgrom1995}. Over time, however, a consensus developed where the
high mass-to-light ratios of dwarf galaxies was due to the presence of a
dark matter halo, which was lacking in GCs
\citep[e.g.,][\textbf{REFS}]{dekel+silk1986}. Since then, the properties
of dwarf galaxies have played an increasingly important role in our
understanding of the clustering of dark matter on small scales {[}e.g.,
sales+2022, bullock+boylan-kolchyn2017{]}.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
and recent star formation. Fig.~\ref{fig:galaxy_images} shows the Large
Magellanic Cloud (LMC) as an example of an irregular, bright dwarf
galaxy, where most stars are in a rotationally-supported thin disk (seen
nearly face-on) with a prominent bar. \textbf{Classical dwarf
spheroidals} occupy intermediate luminosities (
\(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these systems are old,
non-star forming, gas-poor, and spheroidal. All Milky Way satellites
discovered before digital sky surveys are classicals. The 12 classical
dwarf satellites of our Galaxy are (in order of decreasing luminosity)
Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II, Carina, Draco,
Ursa Minor, Canes Venatici I, Sextans I, and Crater II.\footnote{While
  formally the dwarf galaxy names we discuss contain ``dwarf
  spheroidal'' (dSph), e.g.~Sculptor dSph, we omit this suffix for
  brevity.} The \textbf{ultra-faint}s dwarfs (UFDs) occupy the very
faintest magnitude range (\(-7.7 \lesssim M_V\)). These galaxies have
minuscule stellar masses, tend to be more compact, and include the
faintest known galaxies (\textbf{ref}). Altogether, known dwarf galaxies
span more than 15 orders in magnitude, or over 7 decades in stellar
mass. \textbf{Rewrite to include stellar mass ranges as well}

A common definition today for a dwarf galaxy is a stellar system with
evidence for dark matter (high mass-to-light ratios, e.g.~\textbf{REF}).
In contrast, star clusters (like GCs) have no clear evidence for dark
matter.

@@

We only understand a tiny fraction of the universe's composition. The
leading theory of cosmology, \LCDM (Lambda Cold Dark Matter), posits
that the universe is composed of about 68\% dark energy (\(\Lambda\)),
27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers like
  to change definitions of words. \emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion.}
\citep{planckcollaboration+2020}. While the composition of dark matter
and dark energy remains elusive, we know their general properties. Dark
energy causes the acceleration of the expansion of the universe on large
scales. We do not discuss dark energy here---it does not substantially
affect the Local Group today. Dark matter, instead, makes up the vast
majority of mass in galaxies. Typically, galaxies have baryonic to dark
matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies.
In \LCDM, dark matter is assumed to interact only gravitationally. Light
passes through dark matter unimpeded---in this sense dark matter is
transparent. Dark matter is also commonly assumed to be \emph{cold},
i.e.~typical velocities much smaller than the speed of light in the
early universe. If dark matter is cold, then it should condense on all
scales, forming non-linear structures (or \emph{halos}) from the size of
galaxy clusters to smaller than the faintest dwarf galaxies.
Implications of dark matter properties range from cosmological
structural formation, galaxy structure, and galaxy interactions.

\textbf{is this ordered logically?}

\subsection{\texorpdfstring{Dwarf galaxy formation in
\LCDM}{Dwarf galaxy formation in }}\label{dwarf-galaxy-formation-in}

\textbf{In the next section, we consider galaxy formatin in LCDM}

The very early universe was almost featureless. Our earliest
observations of the universe stem from the cosmic microwave background
(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
But tiny perturbations in the CMB, temperature fluctuations of 1 part in
10,000, reveal the underlying seeds of large-scale cosmological
structure. In an expanding universe, gravitational instability makes CDM
overdensities grow and collapse hierarchically onto larger structures.
Initially, baryonic matter was coupled to radiation and resisted
collapse. Dark matter, only influenced by gravity instead, freely
collapsed into the first structures. For mass perturbations both smaller
than the horizon and overdense enough to collapse, these overdensities
of dark matter become self-gravitating, known as \emph{halos}. After
recombination, where electrons combined with atomic nuclei to form
atoms, baryons decoupled from radiation and fell into the dark matter
halos. The densest pockets of baryons later formed the first stars and
galaxies.

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos
\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
Structure formation happens on a wide range of scales. The largest
structures become the cosmic web---composed of voids, filaments, and
sheets---and the smallest non-linear structures (halos, \textbf{define
non-linear}) host dwarf galaxies. Hierarchical assembly is evident
through the large scale structure of the universe, remnants of past
mergers within the Milky Way, and tidal disruption of dwarf galaxies and
their streams around nearby galaxies.

Small-scale structure formation is sensitive to deviations from
\LCDM cosmology \citep[e.g.,][]{bechtol+2022}. One key prediction of
\LCDM is that mass perturbations are expected to exist on all scales,
and are largest on the smallest scales, so we would expect the formation
of halos on all scales. Many alternative models, such as warm dark
matter, may smooth out small-scale features and reduce the abundance of
small halos or change their structure \citep[e.g.,][]{lovell+2014}.
Dwarf galaxies, which occupy the smallest dark matter halos, are
promising windows into small-scale structures. To understand the
evolution of these objects, we need to understand the general
predictions of \LCDM for the properties of dwarf galaxies.

In \LCDM cosmological simulations, dark matter halos are remarkably
self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
observe that the spherically-averaged density profiles \(\rho(r)\) are
universally well described by a two-parameter law:
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density . This profile has shown remarkable success in describing
\LCDM halos across several orders of magnitude in mass. NFW profiles are
\emph{cuspy}, where the density continuously rises like
\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
density profile increases gradually with radius, and at large radii the
density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
shows an example NFW halo.

The total mass of an NFW profile formally diverges, so halo masses are
commonly defined using an overdensity criterion. The virial mass,
\(M_{200}\), is defined as the mass within a radius, \(r_{200}\),
containing a mean enclosed density 200 times\footnote{For the collapse
  of a uniform spherical density, the virialized overdensity would be
  \(\Delta = 18\pi^2\approx 178\) for a critical universe
  \(\Omega_m = 1\). This is commonly rounded to \(\Delta = 200\). While
  this parameter may be closer to \(\Delta \approx 100\) for our
  universe, \(\Delta\) also increases with redshift \citep[using eq. 6
  from][]{bryan+norman1998}.} the critical density of the universe:
\begin{equation}{
M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G
}\end{equation} A standard second parameter of NFW halos is the halo
concentration, \(c=r_{200} / r_s\) which describes how the
characteristic radius scale of the halo compares to the virial radius.
Using this parameter, the scale density is a function of \(c\) alone,
\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
\citep{NFW1996}.

An equivalent, alternative characterization of NFW halos uses their
circular velocity profiles. The circular velocity,
\(\V_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum
\(\V_{\rm max}\) at radius \(r_{\rm max} \approx 2.16258\,r_s\).
\(\V_{\rm max}\) and \(r_{\rm max}\), like \(M_{200}\) and \(c\), fully
specify an NFW halo,

The two parameters of an NFW profile are not independent. Lower mass
dark matter halos often collapse earlier, when the universe was denser.
As a result, low mass subhalos tend to be more concentrated
\citep[e.g.,][]{NFW1997}. The relationship between \(M_{200}\) and c, or
the mass-concentration relation, describes the mean trend of
concentration with mass \citep[e.g.,][]{bullock+2001, ludlow+2016}. The
left panel of Fig.~\ref{fig:smhm} illustrates the present-day
mass-concentration relation in terms of \(r_{\rm max}\) and
\(\V_{\rm max}\). While there is a general trend for concentration to
increase with mass, the concentration values still have substantial
scatter. Other parameters such as the halo spin or shape may affect the
scatter of the mass-concentration relation, but their effect is
typically expected to be small
\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
\caption[Example density profiles]{Density profiles in log density
versus log radius for stars and dark matter in a Sculptor-like galaxy.
The dark matter is more extended and massive than the star across the
entire galaxy.}\label{fig:nfw_density}
\end{figure}

\subsection{\texorpdfstring{Formation of stars within
\LCDM halos}{Formation of stars within halos}}\label{formation-of-stars-within-halos}

The observed abundance of galaxies may be compared with the abundance of
\LCDM halos to derive constraints regarding which galaxies inhabit which
halos. This technique, dubbed ``abundance matching,'' implies a tight
relation between the stellar mass of a galaxy and the mass of the halo
it inhabits \citep{li+white, moster+naab+white2013}.

Fig.~\ref{fig:smhm} shows the stellar mass versus halo mass relation
(SMHM, with halo mass represented by \(\V_{\rm max}\)) predicted by the
\LCDM cosmological hydrodynamical simulations from the the APOSTLE
project \citep{sawala+2016}, which uses the hydrodynamical setup from
the EAGLE project \citep{crain+2015, schaye+2015} to simulate Local
Group analogues in a \LCDM cosmological context \textbf{(db: reorder
these sentences)}. While there is some scatter, the range of predicted
\({\rm v}_{\rm max}\) is fairly narrow across \(\sim 4\) decades in
stellar mass. This figure indicates that the SMHM relation becomes
increasingly steep in the dwarf galaxy regime---many dwarf galaxies are
formed in relatively massive halos. Because lower mass galaxies have
increasingly shallow potential wells, feedback becomes more effective at
removing gas. Re-ionization additionally suppresses late star formation
in the faintest galaxies. As a result, the resulting stellar mass of a
galaxy is highly sensitive to the dark matter mass, especially for faint
dwarf galaxies. If we know the initial stellar mass of a dwarf galaxy,
we also know, at some level, the properties of its dark matter halo.
\textbf{fix this paragraph}

In \LCDM galaxy formation, the majority of mass in dwarf galaxies comes
from Dark Matter. Given the SMHM, Fig.~\ref{fig:nfw_density} shows an
example stellar component for a classical dwarf galaxy with the expected
underlying dark matter halo. Even when the stars are most dense in the
inner regions, the dark matter remains nearly an order of magnitude
higher in density. Stars have a small contribution to the gravitational
structure of dwarf galaxies--instead stars are reasonably approximated
as tracer particles of the underlying dark matter halo. \textbf{add
something about stellar compactness}

Several challenges complicate the SMHM trend, including environment,
assembly history, tidal effects, and the details of galaxy formation. In
particular, effects like ram-pressure stripping (removal of gas in the
dwarf galaxy due to pressure from host's circumgalactic medium) and
tidal removal of gas cause star formation to quench {[}\textbf{REFS}{]}.
Additionally, the time of formation (relative to reionization) can
influence the resulting stellar content \citep{kim+2024}. Finally, tides
influence both the dark matter and stellar mass but in different amounts
\citep[e.g.,][]{PNM2008}. Consequently, tides may reduce the halo mass
more than the stellar mass, adding additional scatter to the SMHM trend,
particularly for low-mass satellites \citep[e.g.,][]{fattahi+2018}.
Understanding the effects of tides on Local Group dwarf galaxies is
essential to determine where and how these galaxies formed in a
cosmological context.

\section{Observing faint features of dwarf
galaxies}\label{observing-faint-features-of-dwarf-galaxies}

Since Local Group dwarfs are nearby, they are resolved into individual
stars, and we can study these galaxies on a star-by-star basis. As a
result, it is possible to measure the 3D velocity and position of a star
if we can measure the its position, distance, line-of-sight (LOS)
velocity, and proper motions. Unfortunately, determining distances and
full 3D velocities is challenging. The most direct measurement of
distance, parallax, requires precise tracking of a star's sky position
across a year. And while line-of-sight (LOS) velocities are relatively
easily determined from spectroscopy, tangential velocities, derived from
proper motions and distances, are much more challenging. Typically,
measuring proper motions requires accurate (much less than arcsecond)
determinations of small changes in a star's position over baselines of
several years to decades. The full 6D position and velocity information
for stars has, until recently, been known for only a handful of stars.

Launched in 2013, \emph{Gaia} is a space-based, all-sky survey telescope
situated at the Sun-Earth L2 Lagrange point
\citep{gaiacollaboration+2016}. \emph{Gaia} has redefined astrometry,
providing photometry, positions, proper motions, and parallaxes for over
1 billion stars \citep{gaiacollaboration+2021}. While \emph{Gaia}
completed its space-based mission in 2025, two forthcoming data releases
are expected.

Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imagining two regions
separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
tiny changes in relative positions of stars across small and large
angles. Combining measurements from multiple epochs across several
years, an absolute all-sky reference frame is derived from which
parallax and proper motions are derived. In addition to astrometry,
\emph{Gaia} measures photometry in the wide \emph{G} band (330-1050nm)
and colours from the blue photometer (BP, 330-680 nm) and red photometer
(RP, 640-1050 nm). \emph{Gaia} additionally provides low resolution
BP-RP spectra and radial velocity measurements of bright stars (of
magnitudes \(G_{\rm RVS} < 16\)) \citep{gaiacollaboration+2016}. For our
work, \emph{Gaia}'s most relevant measurements are \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
  change in right ascension, \(\alpha\), and declination, \(\delta\),
  typically in units of mili-arcsecond (mas) per year.
  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
  effects in \(\alpha\).}

\subsection{The Local Group of
Galaxies}\label{the-local-group-of-galaxies}

The Local Group (LG) is defined as galaxies which are within \(\sim 1\)
Mpc from the Milky Way-Andromeda system \citep[e.g.,][ and references
therein]{mcconnachie2012}. \emph{Gaia} has revolutionized LG studies.
For example, the 6D dynamical measurements and metallicities of MW stars
led to the (re)discovery of past mergers or Milky Way building blocks
like \emph{Gaia}-Sausage Enceladus
\citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
see also][]{mesa+2005}}, out-of-equilibrium structures like the
\emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
helped find and constrain numerous stellar streams
\citep[\citet{bonaca+price-whelan2025}]{ibata+malhan+martin2019}.
Altogether, \emph{Gaia} has revealed the hierarchical formation and
complex, evolving structure of our own Galaxy.

For Milky Way satellites, \emph{Gaia} has enabled well-constrained
orbital analysis and facilitated precise stellar membership
determination. Before, proper motions of dwarf galaxies were ill-known
or undetermined. Few galaxies had precisely measured proper motions,
more recently from Hubble Space Telescope observations
\citep[e.g.,][]{sohn+2017}. \emph{Gaia} allowed the first systematic
determinations of Milky Way satellite proper motions of unprecedented
precision \citep{pace+li2019, MV2020a}. While the proper motion
uncertainty of a typical dwarf member star is often large, by combining
the proper motions of 100s or 1000s of stars from \emph{Gaia}, average
precise proper motion measurements can be determined, sometimes only
limited by \emph{Gaia}'s systematic error floor
\citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a new
dynamical era for MW satellite studies, where we can derive precise
orbits for any satellite, assuming a given MW potential. In addition,
\emph{Gaia} helps establish membership by separate out contaminating MW
foreground stars. By measuring parallaxes and/or proper motions, many
background and foreground stars can be classified as non-members
\citep[e.g.,][\citet{jensen+2024}]{battaglia+2022}.

@@

@@

Today, we know that the Milky Way system is teeming with satellites.
Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
including dwarf galaxies, globular clusters, and ambiguous systems,
whose nature as a galaxy or star cluster remains uncertain. Advances in
observational techniques has accelerated progress in our understanding
of dwarf galaxies and introduced new questions. Deep digital photometric
surveys have more than quadrupled the number of known Milky Way dwarf
galaxies since 2005 \citep{simon2019}. Upcoming surveys, like the Vera
Rubin Observatory's Legacy Survey for Space and Time
\citep[LSST,][]{ivezic+2019}, will likely identify even fainter dwarf
galaxies. In addition, large aperture multi-object spectrographs have
revealed the detailed and complex inner chemodynamical structure of
dwarf galaxies \textbf{REFS}. Beyond precise structural and kinematic
properties of dwarf galaxies, modern observations allow for the
separation of multiple stellar populations, detailed constraints on the
dark matter density profiles, and hints of tidal disruption or stellar
halos.

Of local dwarfs, the classical systems still remain the most studied
galaxies with the best constrained parameters. These dwarfs also contain
large numbers of bright (giant) stars which can be followed up
spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a result,
the determination of fundamental properties such as the position, size,
orientation, distances, proper motions, line-of-sight (LOS) velocities,
and velocity dispersions \(\sigma_\V\), are all relatively well
constrained today.

Observations of dwarf galaxies have been the origin of several disputes
or \emph{small-scale} problems for
\LCDM [see reviews by @bullock+boylan-kolchin2017; sales+2022]. For
example, the mismatch between the expected number of dwarf galaxies from
simulations and the observed abundance has been known as the
\emph{missing satellites problem}. Additionally, a number of
observations suggested that some dwarf galaxies, although not all,
possess dark matter ``cores,''
\citep[e.g.,][]{moore1994, adams+2014, oh+2015, walker+penarrubia2011, read+walker+steger2019},\footnote{In
  detail, gas-phase rotation curves are better able to differentiate
  between cores and cusps, whereas stellar kinematics is less
  constraining.} contrary to the expectation from \LCDM of ``cuspy''
inter dark matter profiles \citep{NFW1996, NFW1997}. As a result,
alternative forms of dark matter have been advocated as solutions, such
as Warm or Self-Interacting Dark Matter.
